### 2023-02-26 calculus.tex
Describe marking metatheorems, start product extension section

@@

@@

\subsection{Correctness}
\begin{theorem}[name=Marking Totality]
  Any syntactically well-formed program may be marked for errors.
\end{theorem}

In the next section, we will formally develop

@@

  \item \textbf{Marked expressions} (\cref{sec:calculus-mexp}), which mirror the structure of
    unmarked expressions but are extended with \textbf{error markings}. Every marked expression
    corresponds to an unmarked expression by \emph{mark erasure}.

  \item \textbf{Mark insertion} (\cref{sec:calculus-marking}), which \textbf{marks} any unmarked expression into a marked
    expression, inserting error markings where appropriate.
\end{itemize}
%
That is, we extend the ordinary typing semantics of a language with a secondary ``marked'' language
and define a mapping from ``unmarked'' programs to ``marked'' programs. Note that while the
semantics that follow concern a gradually typed language, a non-gradual system may be recovered
straightforwardly by removing the unknown type and replacing type consistency with equality.

In \cref{sec:calculus-correctness}, we discuss the metatheory that ensures the system---and any such
error localization semantics---to be correct. We follow then by extending the calculus with support
for product types. In the interest of space, this overview focuses on ``interesting'' cases. All
rules and theorems may be found in the supplementary appendix, alongside a complete mechanization
(including the extension with products) in the Agda proof assistant, which is discussed in
\cref{sec:calculus-agda}.

@@

\subsection{Mark Insertion}
\label{sec:calculus-marking}

The central component of the calculus is \emph{mark insertion}, a bidirectional system given by the
following mutually defined judgments. The synthetic insertion judgment
$\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$ establishes that under the context $\Gamma$, the
unmarked expression $\EMV$ is ``marked into'' the marked expression $\ECMV$, which synthesizes type
$\TMV$. Analogously, the analytic insertion judgment $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$
states that the unmarked expression $\EMV$ is marked into $\ECMV$, which analyses against type
$\TMV$.

@@

The rules for mark insertion are given in \Cref{fig:calculus-mark-insertion}. Observe that they
mirror the typing rules for marked expressions closely: all possible outcomes for fallible premises
are handled and markings are inserted as needed. For example, ISVar and ISUnbound together handle
variable usages. The former yields the same variable in the marked language when it is bound. The
latter marks the variable as unbound if it is not found in the context.

We again find two rules governing the marking of function application. ISAp1 corresponds to MSAp1 as
the ``success case'', recursively marking $\EMV_1$ and $\EMV_2$. Accordingly, ISAp2 parallels MSAp2
as the ``error case''. Similarly, ISIf and ISInconsistentBranches correspond to MSIf and
MSInconsistentBranches, and IALam1, IALam2, and IALam3 correspond to MSLam1, MSLam2, and MSLam3.

\subsection{Correctness}
\label{sec:calculus-correctness}

% Totality

% Well-formedness conditions

% Well-typed expressions have no resulting marks

\subsection{Adding Products}
\label{sec:calculus-products}

@@

>>>>

% TODO standardize naming: marking vs mark insertion, marks vs error marks vs markings
% TODO uniform style for referencing rules

@@

% TODO examples of this ad hoc stuff

@@

@@

  \item \textbf{Marked expressions} (\cref{sec:calculus-mexp}), which mirror the structure of
    unmarked expressions but are extended with \textbf{error marks}. Every marked expression
    corresponds to an unmarked expression by \emph{mark erasure}.
    % TODO more intuition for what marked expressions are?

  \item \textbf{Marking} (\cref{sec:calculus-marking}), which \textbf{marks} any unmarked expression into a marked
    expression, inserting error markings where appropriate.
\end{itemize}
%
That is, we extend the ordinary typing semantics of a language with a secondary ``marked'' language
and define a mapping from ``unmarked'' programs to ``marked'' programs. Note that while the
semantics that follow concern a gradually typed language, a non-gradual system may be recovered
straightforwardly by removing the unknown type and replacing type consistency with equality.

Key to ensuring correctness of the system---and, indeed, any such error localization semantics---is
a metatheorem, \emph{marking totality}, which guides the construction of rules:
%
\begin{theorem}[name=Marking Totality]
  Any syntactically well-formed unmarked expression may be marked into a marked expression.
\end{theorem}
%
We will give a more concrete statement of the metatheorem after formally introducing the components
of the system.

In \cref{sec:calculus-products}, we follow with an extension of the calculus to support binary
product types. In the interest of space, this overview focuses on ``interesting'' cases. All rules
and theorems may be found in the supplementary appendix, alongside a complete mechanization
(including the extension with binary products) in the Agda proof assistant, which is discussed in
\cref{sec:calculus-agda}.

@@

\subsection{Marking}
\label{sec:calculus-marking}

The central component of the calculus is \emph{marking}, a bidirectional system given by the
following mutually defined judgments. The synthetic insertion judgment
$\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$ establishes that, under the context $\Gamma$, the
unmarked expression $\EMV$ is ``marked into'' the marked expression $\ECMV$, which synthesizes type
$\TMV$. Analogously, the analytic insertion judgment $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$
states that the unmarked expression $\EMV$ is marked into $\ECMV$, which analyses against type
$\TMV$.

@@

The rules for mark insertion are given in \cref{fig:calculus-mark-insertion}. Observe that they
mirror the typing rules for marked expressions closely: all possible outcomes for fallible premises
are handled and markings are inserted as needed. For example, ISVar and ISUnbound together handle
variable usages. The former yields the same variable in the marked language when it is bound. The
latter marks the variable as unbound if it is not found in the context.

We again find two rules governing the marking of function application. ISAp1 corresponds to MSAp1 as
the ``success case'', recursively marking $\EMV_1$ and $\EMV_2$. Accordingly, ISAp2 parallels MSAp2
as the ``error case''. Similarly, ISIf and ISInconsistentBranches correspond to MSIf and
MSInconsistentBranches, and IALam1, IALam2, and IALam3 correspond to MSLam1, MSLam2, and MSLam3.
The rules IAInconsistentTypes and IASubsume, of course, act in parallel to MAInconsistentTypes and
MASubsume.

We may now formally restate the critical metatheorem of marking totality:
%
\begin{theorem}[name=Marking Totality] \
  \label{thm:calculus-marking-totality}
  \begin{enumerate}
    \item For all $\ctx$ and $\EMV$, there exist $\ECMV$ and $\TMV$ such that
      $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$.
  \item For all $\ctx$, $\EMV$, and $\TMV$, there exists $\ECMV$ such that
    $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$.
  \end{enumerate}
\end{theorem}

Additionally, it is important that marking preserves the original syntactic structure modulo error
marks; it would be trivial to satisfy totality by marking all expressions into numbers, for example.
This property is ensured by the following well-formedness metatheorem:
%
\begin{theorem}[name=Marking Well-Formedness] \
  \label{thm:calculus-marking-well-formedness}
  \begin{enumerate}
    \item If $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$, then $\ctxSynType{\ctx}{\ECMV}{\TMV}$ and
      $\erasesTo{\ECMV}{\EMV}$.
    \item If $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$, then $\ctxAnaType{\ctx}{\ECMV}{\TMV}$ and
      $\erasesTo{\ECMV}{\EMV}$.
  \end{enumerate}
  % TODO split this into two theorems?
  % TODO not really sure how to intuitively describe the part about types matching up
\end{theorem}
%
In other words, if an unmarked expression is marked into some marked expression, erasing error marks
from the marked expression gives the original unmarked expression. To ensure that we place error
marks in the correct spots, we furthermore require the following theorem:
%
\begin{theorem}[name=Marking of Well-Typed/Ill-Typed Expressions] \ % TODO better name for this
  \label{thm:calculus-marking-well-ill-typed}
  \begin{enumerate}
    \item \begin{enumerate}
        \item If $\ctxSynType{\ctx}{\EMV}{\TMV}$, then $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$
          such that $\imageIs{\EMV}{\ECMV}$.
        \item If $\ctxAnaType{\ctx}{\EMV}{\TMV}$, then $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$
          such that $\imageIs{\EMV}{\ECMV}$.
      \end{enumerate}

    \item \begin{enumerate}
        \item If there does not exist $\TMV$ such that $\ctxSynType{\ctx}{\EMV}{\TMV}$, then for all
          $\ECMV$ and $\TMV'$ such that $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV'}$, we have
          $\imageIsNot{\EMV}{\ECMV}$.
        \item If there does not exist $\TMV$ such that $\ctxAnaType{\ctx}{\EMV}{\TMV}$, then for all
          $\ECMV$ and $\TMV'$ such that $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV'}$, we have
          $\imageIsNot{\EMV}{\ECMV}$.
      \end{enumerate}
  \end{enumerate}
\end{theorem}
%
That is, since error marks indicate type errors in the program, a well-typed program should, of
course, have no error marks. The converse---that marks are inserted into ill-typed programs---is
also true. We have chosen to formulate the absence of error marks through an inclusion mapping of
unmarked expressions into marked expressions; alternatively, a predicate over marked expressions
could be used. % TODO latter is easier to do in Agda, but maybe it is weaker?

Finally, no less importantly, marking is deterministic. This is given by
\cref{thm:calculus-marking-unicity}.
%
\begin{theorem}[name=Marking Unicity] \
  \label{thm:calculus-marking-unicity}
  \begin{enumerate}
    \item If $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV_1}{\TMV_1}$ and
      $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV_2}{\TMV_2}$, then $\ECMV_1 = \ECMV_2$ and $\TMV_1 =
      \TMV_2$.
    \item If $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV_1}{\TMV_1}$ and
      $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV_2}{\TMV_2}$, then $\ECMV_1 = \ECMV_2$ and $\TMV_1 =
      \TMV_2$.
  \end{enumerate}
\end{theorem}
%
Together with totality, marking may be considered and implemented as a total function.

\subsection{Adding Binary Products}
\label{sec:calculus-products}

We now demonstrate how one might extend or adapt the marking calculus for new typing features by
adding binary product types. First, we extend in \cref{fig:calculus-products-judgments} the syntax
of types and notions of consistency and join. As with arrow types, we additionally define an
auxiliary notion of \emph{matched binary product types}: the judgment
$\matchedPair{\TMV}{\TMV_1}{\TMV_2}$ establishes that $\TMV$ may be considered the binary product
type $\TPair{\TMV_1}{\TMV_2}$.

\input{figs/calculus-products-judgments}

The unmarked language with a pair constructor and projection operators.
\Cref{fig:calculus-products-uexp} gives their syntax and typing rules. Similarly to application,
\textsc{USProjL} and \textsc{USProjR} employ the matched binary product type; a projection
$\EProjL{\EMV}$ or $\EProjR{\EMV}$ is then well-typed if the $\EMV$ may be considered to be a pair.

\input{figs/calculus-products-uexp}

\input{figs/calculus-products-mexp}

\input{figs/calculus-products-marking}

@@

\subsection{Implementation in Hazel}

\subsection{Reified Hazelnut Action Semantics}
